---
title: "Sampling La Plata"
author: "Einav Grinberg"
date: "10/25/2021"
output: bookdown::html_document2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(sf)
library(raster)
library(plotrix)  
library(sfsmisc)
library(mapview)
library(rgdal)
library(knitr)
setwd("C:/Users/einav/Desktop/Thesis data/La Plata")
```

# Simple Random Spatial Sampling 

## La Plata sample size calculation

The following formula can help decide the most suitable size for a sample:

\begin{equation} 
  n_0 = \frac{Z^2p(1-p)}{e^2} (\#eq:sample)
\end{equation} 

Where $n_0$ is the sample size, $Z^2$ is the confidence level Z-score ([can be found in this table](https://www.sjsu.edu/faculty/gerstman/StatPrimer/z-two-tails.pdf)), $p$ is the estimated proportion of variability in the population and $e^2$ is the margin of error, a.k.a. a confidence interval.

**Usually with a large population where there is no knowledge about the proportion of variability in the population -> $p=0.5$ (the maximum variability).

La Plata suitable sample size with a 5% confidence interval:

\begin{equation} 
  n_0 = \frac{1.96^2*0.5*(1-0.5)}{0.05^2} = 384.16 (\#eq:sample1)
\end{equation} 

Therefore the most suitable sample to collect for La Plata would be 384 building rooftops.

In this analysis two sets of random spatial samples have been drawn from the city of La Plata:

1. A sample of 100 building rooftops
2. A sample of 300 building rooftops
3. And both samples combined

The samples were created using the **Vector Research Tools** -> **Random points inside a polygon** in QGIS^[https://www.qgis.org/en/site/]. Then the building rooftops were digitized in QGIS using Google Satellite Hybrid^[https://qms.nextgis.com/geoservices/1135/], a Tile Map Service (TMS) layer.

For every sample the rooftop area $m^2$, the mean global horizontal irradiation $(\frac{kWh}{m^2})$, the usable solar radiation $(kWh)$ and renewable electricity production $(kWh)$ were calculated. 

*Buildings rooftops area that are equal and under 30 $m^2$ are defined as 0. Some of the sample points were computed on non built-up areas/roads/parks etc., therefore they were given a 0 to include a density factor in the calculation.*

```{r random sampling, warning=FALSE, echo=FALSE}
laplata_poly <- read_sf("polygon_3.shp")
laplata_poly <- st_transform(laplata_poly, "+proj=merc +lon_0=0 +k=1 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs")

sample1 <- read_sf("sample100.shp")
sample1 <- st_transform(sample1, "+proj=merc +lon_0=0 +k=1 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs")
#change to 0 above and equal 30 sqm
sample1$elec_prod[sample1$area <= 30] <- 0
#calculate mega watt per hour column
sample1$elec_prod_mwh <- sample1$elec_prod/1000

head(sample1)
summary(sample1)

sample2 <- read_sf("sample300.shp")
sample2 <- st_transform(sample2, "+proj=merc +lon_0=0 +k=1 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs")
#change to 0 above and equal 30 sqm
sample2$elec_prod[sample2$area <= 30] <- 0
sample2$elec_prod_mwh <- sample2$elec_prod/1000

head(sample2)
summary(sample2)

mapview(laplata_poly,  map.types =  "CartoDB.VoyagerLabelsUnder", alpha.regions = 0, legend=F, color="red", lwd=2) + mapview(sample1, col.regions="blue") + mapview(sample2, col.regions="green")
```

## Sample 1 computations

Calculation of the sample mean is represented by $\overline{y}$.
Calculation of the sample variance is represented by $s$.

```{r sample 1 c, warning=FALSE, echo=FALSE}
boxplot(sample1$elec_prod_mwh, horizontal=TRUE, main="Sample 1", xlab = "Renewable Electricity Production Potential (mWh)")

n1 <- as.numeric(nrow(sample1))
y1 <- mean(sample1$elec_prod_mwh)
paste("The mean of sample 1: ", round(y1,2), "(mWh)")
N_all <- 2482 #all buildings in the city
s1 <- var(sample1$elec_prod_mwh)
paste("The variance of sample 1: ", round(s1,2), "(mWh)")
```
The following equation calculates the unbiased variance of the estimator $\overline{y}$: 

\begin{equation} 
  \hat{var}(\overline{y})= (\frac{N-n}{N})*(\frac{s^2}{n}) (\#eq:unvar)
\end{equation} 

* $N$ is the total population size -> the number of all the buildings in La Plata (2,482). 
* $n$ is the sample size -> 100. 
* $s^2$ is the sample variance. 

The following equation calculates the estimated standard error of the estimator $\overline{y}$: 

\begin{equation} 
  SEM = \sqrt{\hat{var}(\overline{y})} (\#eq:se)
\end{equation} 

```{r variance, warning=FALSE, echo=FALSE}
var_un1 <- ((N_all-n1)/N_all)*(s1/n1)
paste("The variance of the sample mean: ", round(var_un1,2) , "(mWh)")

ese1 = sqrt(var_un1) #estimated standard error of variance
paste("The estimated standard error of the sample mean: ", round(ese1,2), "(mWh)")
```

The following equation calculates an unbiased estimator of the population total $\hat{t}$: 
\begin{equation} 
  \hat{t} = N*{\overline{y}} (\#eq:t)
\end{equation} 

The following equation calculates the unbiased variance of the estimator $\hat{t}$: 

\begin{equation} 
  \hat{var}(\hat{t})= N^2*\hat{var}(\overline{y}) (\#eq:tv)
\end{equation} 

The following equation calculates the estimated standard error of the estimator $\hat{t}$: 
\begin{equation} 
  SEM = \sqrt{\hat{var}(\hat{t})} (\#eq:set)
\end{equation} 

```{r total, warning=FALSE, echo=FALSE}
t_un1 = N_all*y1 
paste("The estimation of the renewable electricity production potential by all the buildings in the city:", round(t_un1,2), "(mWh)")

t_var_un1 = (N_all^2) * var_un1 #unbiased variance with the estimate of the total
paste("The variance of the estimated total:", round(t_var_un1,2), "(mWh)")

ese_t1 = sqrt(t_var_un1) #estimated standard error of total
paste("The estimated standard error of the total:", round(ese_t1,2), "(mWh)")
```

```{r CI 1, warning=FALSE ,echo=FALSE}
#confidence interval with 95% 
CI1 <- t_un1 + qt(c(0.05,0.95), df=99) * ese_t1
paste("The 95% confidence interval estimation for sample 1 is: ", "(",round(CI1[1],2)," (mWh), ",round(CI1[2],2)," (mwh))", sep = "")
```

## Sample 2 computations

Calculation of the sample mean is represented by $\overline{y}$.
Calculation of the sample variance is represented by $s$.

```{r sample 2 c, warning=FALSE, echo=FALSE}
boxplot(sample2$elec_prod_mwh, horizontal=TRUE, main="Sample 2", xlab = "Renewable Electricity Production Potential (mWh)")

n2 <- as.numeric(nrow(sample2))
y2 <- mean(sample2$elec_prod_mwh)
paste("The mean of sample 2: ", round(y2,2), "(mWh)")
N_all <- 2482 #all buildings in the city
s2 <- var(sample2$elec_prod_mwh)
paste("The variance of sample 2: ", round(s2,2), "(mWh)")
```

Equation \@ref(eq:unvar) calculates the unbiased variance of the estimator $\overline{y}$. 

Equation \@ref(eq:se) calculates the estimated standard error of the estimator $\overline{y}$.

```{r variance 2, warning=FALSE, echo=FALSE}
var_un2 <- ((N_all-n2)/N_all)*(s2/n2)
paste("The variance of the sample mean: ", round(var_un2,2), "(mWh)")

ese2 = sqrt(var_un2) #estimated standard error of variance
paste("The estimated standard error of the sample mean: ", round(ese2,2), "(mWh)")
```

Equation \@ref(eq:t) calculates an unbiased estimator of the population total $\hat{t}$.

Equation \@ref(eq:tv) calculates the unbiased variance of the estimator $\hat{t}$.

Equation \@ref(eq:set) calculates the estimated standard error of the estimator $\hat{t}$.

```{r total 2, warning=FALSE, echo=FALSE}
t_un2 = N_all*y2 
paste("The estimation of the renewable electricity production potential by all the buildings in the city:", round(t_un2,2), "(mWh)")

t_var_un2 = (N_all^2) * var_un2 #unbiased variance with the estimate of the total
paste("The variance of the estimated total:", round(t_var_un2,2), "(mWh)")

ese_t2 = sqrt(t_var_un2) #estimated standard error of total
paste("The estimated standard error of the total:", round(ese_t2,2), "(mWh)")
```

```{r CI 2, warning=FALSE ,echo=FALSE}
#confidence interval with 95% 
CI2 <- t_un2 + qt(c(0.05,0.95), df=299) * ese_t2
paste("The 95% confidence interval estimation for sample 2 is: ", "(",round(CI2[1],2)," (mWh), ",round(CI2[2],2)," (mwh))", sep = "")
```

## Sample 1 and 2 combined computations 

The following calculations are for both sample 1 and sample 2 together, resulting in a total sample of 400 building rooftops. We will call this sample, sample 3. 

Calculation of the sample mean is represented by $\overline{y}$.
Calculation of the sample variance is represented by $s$.
```{r both, warning=FALSE, echo=FALSE}
#both samples
sample3 <- rbind(sample1, sample2)

n3 <- as.numeric(nrow(sample3))
y3 <- mean(sample3$elec_prod_mwh)
paste("The mean of sample 3: ", round(y3,2), "(mWh)")
N_all <- 2482 #all buildings in the city
s3 <- var(sample3$elec_prod_mwh)
paste("The variance of sample 3: ", round(s3,2), "(mWh)")
```

Equation \@ref(eq:unvar) calculates the unbiased variance of the estimator $\overline{y}$. 

Equation \@ref(eq:se) calculates the estimated standard error of the estimator $\overline{y}$.

```{r variance 3, warning=FALSE, echo=FALSE}
var_un3 <- ((N_all-n3)/N_all)*(s3/n3)
paste("The variance of the sample mean: ", round(var_un3,2), "(mWh)")

ese3 = sqrt(var_un3) #estimated standard error of variance
paste("The estimated standard error of the sample mean: ", round(ese3,2), "(mWh)")
```

Equation \@ref(eq:t) calculates an unbiased estimator of the population total $\hat{t}$.

Equation \@ref(eq:tv) calculates the unbiased variance of the estimator $\hat{t}$.

Equation \@ref(eq:set) calculates the estimated standard error of the estimator $\hat{t}$.

```{r total 3, warning=FALSE, echo=FALSE}
t_un3 = N_all*y3
paste("The estimation of the renewable electricity production potential by all the buildings in the city:", round(t_un3,2), "(mWh)")

t_var_un3 = (N_all^2) * var_un3 #unbiased variance with the estimate of the total
paste("The variance of the estimated total:", round(t_var_un3,2), "(mWh)")

ese_t3 = sqrt(t_var_un3) #estimated standard error of total
paste("The estimated standard error of the total:", round(ese_t3,2), "(mWh)")
```

```{r CI 3, warning=FALSE ,echo=FALSE}
#confidence interval with 95% 
CI3 <- t_un3 + qt(c(0.05,0.95), df=399) * ese_t3
paste("The 95% confidence interval estimation for sample 3 is: ", "(",round(CI3[1],2)," (mWh), ",round(CI3[2],2)," (mwh))", sep = "")
```

# Stratified Random Spatial Sampling 

## Equal sample size stratification

La Plata is divided to 2 stratas based on satellite imagery provided by the Copernicus Land Monitoring Service global maps of land cover & cover changes and related surface area statistics^[https://lcviewer.vito.be/about]. 

Strata 1 represents built up area in the city and Strata 2 represents non built up area.

For each strata 60 random points were computed and then 20 building rooftops were digitized using the same methods as in chapter \@ref(simple-random-spatial-sampling).

```{r stratified sampling, warning=FALSE, echo=FALSE}
landuse <- raster("laplata_lubu.tif")

bu <- read_sf("strata_bu.shp")
bu <- st_transform(bu, "+proj=merc +lon_0=0 +k=1 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs")
bu$landuse <- "Built up"
nonbu <- read_sf("strata_nonbu.shp")
nonbu <- st_transform(nonbu, "+proj=merc +lon_0=0 +k=1 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs")
nonbu$landuse <- "Non built up"

#combine stratas and change to 0 above and equal 30 sqm
strat <- rbind(bu, nonbu)
strat$elec_prod[strat$area <= 30] <- 0
strat$elec_prod_mwh <- strat$elec_prod/1000

head(strat)
summary(strat)

mapview(landuse,  map.types =  "CartoDB.VoyagerLabelsUnder", alpha.regions = 0.15, legend=F, col.regions=list("yellow","black")) + mapview(strat, zcol="landuse")
```

After stratification, the stratas are combined to one sample and the computations are the same as for random sampling.

```{r strat c, warning=FALSE,echo=FALSE}
boxplot(strat$elec_prod_mwh ~ strat$landuse, main="Samples from La Plata Land Use Classification", ylab = "Renewable Electricity Production Potential (mWh)", xlab = "Strata by Land Use",cex.axis = 0.7)

nstrat <- as.numeric(nrow(strat))
# calculate means for each group
aggregate(data=strat, elec_prod_mwh~landuse, FUN=mean)

ystrat <- mean(strat$elec_prod_mwh)
paste("The mean of the stratfied sample is: ", round(ystrat,2), "(mWh)")
N_all <- 2482 #all buildings in the city
sstrat <- var(strat$elec_prod_mwh)
paste("The variance of the stratfied sample is: ", round(sstrat,2), "(mWh)")
```

Equation \@ref(eq:unvar) calculates the unbiased variance of the estimator $\overline{y}$. 

Equation \@ref(eq:se) calculates the estimated standard error of the estimator $\overline{y}$.

```{r variance strat, warning=FALSE, echo=FALSE}
var_unstrat <- ((N_all-nstrat)/N_all)*(sstrat/nstrat)
paste("The variance of the sample mean: ", round(var_unstrat,2), "(mWh)")

esestrat = sqrt(var_unstrat) #estimated standard error of variance
paste("The estimated standard error of the sample mean: ", round(esestrat,2), "(mWh)")
```

Equation \@ref(eq:t) calculates an unbiased estimator of the population total $\hat{t}$.

Equation \@ref(eq:tv) calculates the unbiased variance of the estimator $\hat{t}$.

Equation \@ref(eq:set) calculates the estimated standard error of the estimator $\hat{t}$.

```{r total strat, warning=FALSE, echo=FALSE}
t_unstrat = N_all*ystrat
paste("The estimation of the renewable electricity production potential by all the buildings in the city:", round(t_unstrat,2), "(mWh)")

t_var_strat = (N_all^2) * var_unstrat #unbiased variance with the estimate of the total
paste("The variance of the estimated total:", round(t_var_strat,2), "(mWh)")

ese_tstrat = sqrt(t_var_strat) #estimated standard error of total
paste("The estimated standard error of the total:", round(ese_tstrat,2), "(mWh)")
```

```{r CI strat, warning=FALSE ,echo=FALSE}
#confidence interval with 95% 
CIstrat <- t_unstrat + qt(c(0.05,0.95), df=119) * ese_tstrat
paste("The 95% confidence interval estimation for the stratified sample is: ", "(",round(CIstrat[1],2)," (mWh), ",round(CIstrat[2],2)," (mwh))", sep = "")
```

## Optimal allocation stratification

When dividing the city of La Plata into built up and non built up area, the sample calculations from the built up area present more variable strata. 
The optimum scheme allocates larger sample size to the more variable strata and smaller sample size to the more difficult-to-sample strata.^[Thompson, S. (2012). Sampling. Hoboken, N.J.: Wiley.p.147]

Therefore, 2 new stratas are computed - 

1. Optimally allocated strata in the built up area - a sample of 95 building rooftops
2. Optimally allocated strata in the non built up area - a sample of 5 building rooftops.

```{r stratified allocation sampling, warning=FALSE, echo=FALSE}
landuse <- raster("laplata_lubu.tif")

bu_prop <- read_sf("strata_prop_bu.shp")
bu_prop <- st_transform(bu_prop, "+proj=merc +lon_0=0 +k=1 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs")
bu_prop$landuse <- "Built up"
nonbu_prop <- read_sf("strata_prop_nonbu.shp")
nonbu_prop <- st_transform(nonbu_prop, "+proj=merc +lon_0=0 +k=1 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs")
nonbu_prop$landuse <- "Non built up"

#combine stratas and change to 0 above and equal 30 sqm
strat_prop <- rbind(bu_prop, nonbu_prop)
strat_prop$elec_prod[strat_prop$area <= 30] <- 0
strat_prop$elec_prod_mwh <- strat_prop$elec_prod/1000

head(strat_prop)
summary(strat_prop)

mapview(landuse,  map.types =  "CartoDB.VoyagerLabelsUnder", alpha.regions = 0.15, legend=F, col.regions=list("yellow","black")) + mapview(strat_prop, zcol="landuse")
```

After stratification, the stratas are combined to one sample and the computations are the same as for random sampling.

```{r strat_prop c, warning=FALSE,echo=FALSE}
boxplot(strat_prop$elec_prod_mwh ~ strat_prop$landuse, main="Samples from La Plata Land Use Classification", ylab = "Renewable Electricity Production Potential (mWh)", xlab = "Proportional Allocated Strata by Land Use",cex.axis = 0.7)

nstrat_prop <- as.numeric(nrow(strat_prop))
# calculate means for each group
aggregate(data=strat_prop, elec_prod_mwh~landuse, FUN=mean)

ystrat_prop <- mean(strat_prop$elec_prod_mwh)
paste("The mean of the stratfied sample is: ", round(ystrat_prop,2), "(mWh)")
N_all <- 2482 #all buildings in the city
sstrat_prop <- var(strat_prop$elec_prod_mwh)
paste("The variance of the stratfied sample is: ", round(sstrat_prop,2), "(mWh)")
```
Equation \@ref(eq:unvar) calculates the unbiased variance of the estimator $\overline{y}$. 

Equation \@ref(eq:se) calculates the estimated standard error of the estimator $\overline{y}$.

```{r variance strat_prop, warning=FALSE, echo=FALSE}
var_unstrat_prop <- ((N_all-nstrat_prop)/N_all)*(sstrat_prop/nstrat_prop)
paste("The variance of the sample mean: ", round(var_unstrat,2), "(mWh)")

esestrat_prop = sqrt(var_unstrat_prop) #estimated standard error of variance
paste("The estimated standard error of the sample mean: ", round(esestrat_prop,2), "(mWh)")
```

Equation \@ref(eq:t) calculates an unbiased estimator of the population total $\hat{t}$.

Equation \@ref(eq:tv) calculates the unbiased variance of the estimator $\hat{t}$.

Equation \@ref(eq:set) calculates the estimated standard error of the estimator $\hat{t}$.

```{r total strat_prop, warning=FALSE, echo=FALSE}
t_unstrat_prop = N_all*ystrat_prop
paste("The estimation of the renewable electricity production potential by all the buildings in the city:", round(t_unstrat_prop,2), "(mWh)")

t_var_strat_prop = (N_all^2) * var_unstrat_prop #unbiased variance with the estimate of the total
paste("The variance of the estimated total:", round(t_var_strat_prop,2), "(mWh)")

ese_tstrat_prop = sqrt(t_var_strat_prop) #estimated standard error of total
paste("The estimated standard error of the total:", round(ese_tstrat_prop,2), "(mWh)")
```

```{r CI strat_prop, warning=FALSE ,echo=FALSE}
#confidence interval with 95% 
CIstrat_prop <- t_unstrat_prop + qt(c(0.05,0.95), df=99) * ese_tstrat_prop
paste("The 95% confidence interval estimation for the stratified sample is: ", "(",round(CIstrat_prop[1],2)," (mWh), ",round(CIstrat_prop[2],2)," (mwh))", sep = "")
```


# 95% Confidence Intervals

```{r CI plot, echo=FALSE, warning=FALSE}
table <- data.frame(sample=c("Sample 1 - 100", "Sample 2 - 300", "Both Samples - 400", "Stratified Sample - 120", "Allocated Stratified Sample - 100"),
                    mean=c(y1, y2, y3, ystrat, ystrat_prop),
                    variance_of_mean=c(var_un1, var_un2, var_un3, var_unstrat, var_unstrat_prop),
                    standard_error_mean=c(ese1,ese2,ese3,esestrat, esestrat_prop), total_estimation=c(t_un1, t_un2, t_un3, t_unstrat, t_unstrat_prop), variance_of_total=c(t_var_un1, t_var_un2, t_var_un3, t_var_strat, t_var_strat_prop), standard_error_total=c(ese_t1, ese_t2, ese_t3, ese_tstrat, ese_tstrat_prop), CI_low=c(CI1[1],CI2[1],CI3[1],CIstrat[1],CIstrat_prop[1]), CI_up=c(CI1[2],CI2[2],CI3[2],CIstrat[2],CIstrat_prop[2]))


plotCI(y = table$total_estimation,               # plotrix plot with confidence intervals
       x = 1:5, li = table$CI_low,
       ui = table$CI_up ,scol="red", xaxt="n", yaxt="n",
       xlab="", ylab="Renewable Electricity (mWh)", main="Confidence Intervals", xlim=c(1,5))
#text(x = totals$sample,y = totals$value, totals$value, pos = 3,srt=45)
axis(side=1,at=1:5,  ## add custom x-axis
     label=c("sample 1 - 100", "sample 2 - 300", "both samples - 400","stratified sample - 120", "allocated stratified sample - 100"), cex.axis = 0.7)
eaxis(2, at.small = F, cex.axis = 0.7)
```

# Summarizing Table

```{r table, warning=FALSE,echo=FALSE}

kable(table, digits = 2, format.args = list(big.mark = ",", scientific = FALSE), col.names = c('Sample', 'Mean', 'Var of Mean', 'SEM', 'Total Estimation', 'Var of Total', 'SET', 'Lower CI', 'Upper CI'), algin="lccccccccc", caption = "Summary of Computation Results (mWh)")
```
